---
title: "Gene expression"
author: "Rahmat"
date: "2024-11-07"
output: html_document
---

## Gene Expression analysis
```{r}
setwd("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts")
```

# IMPORTING COUNTS
```{r}
SRR1313090 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR1313090.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)
head(SRR1313090)
```
```{r}
SRR1313091 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR1313091.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)
head(SRR1313091)
```
```{r}
SRR1313092 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR1313092.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)
head(SRR1313092)
```
```{r}
SRR1313093 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR1313093.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)
head(SRR1313093)
```
```{r}
SRR1313148 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR1313148.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)
head(SRR1313148)
```
```{r}
SRR1313149 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR1313149.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)
head(SRR1313149)
```
```{r}
SRR1313150 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR1313150.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)
head(SRR1313150)
```
```{r}
SRR1313151 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR1313151.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)
```
```{r}
SRR1313204 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR1313204.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)
head(SRR1313204)
```
```{r}
SRR1313205 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR1313205.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)

```
```{r}
SRR1313206 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR1313206.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)
head(SRR1313206)
```
```{r}
SRR1313208 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR1313208.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)
head(SRR1313208)
```
```{r}
SRR13873745 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR13873745.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)

```
```{r}
SRR13873746 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR13873746.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)

```
```{r}
SRR13873747 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR13873747.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)

```
```{r}
SRR13873748 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR13873748.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)

```
```{r}
SRR22891574 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR22891574.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)

```
```{r}
SRR22891592 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR22891592.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)

```
```{r}
SRR22891594 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR22891594.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)

```
```{r}
SRR22891598 <- read.table("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Counts/SRR22891598.tabular", header= TRUE, row.names = 1, stringsAsFactors = TRUE)

```

# Merging counts data
```{r}
count_data <- cbind(SRR1313090,SRR1313091,SRR1313092,SRR1313093,SRR13873745,SRR13873746,SRR13873747,SRR13873748,SRR1313148,SRR1313149,SRR1313150,SRR1313151,SRR1313204,SRR1313205,SRR1313206,SRR1313208,SRR22891574,SRR22891592,SRR22891594,
SRR22891598)
head(count_data)
```
```{r}
count_data <- count_data[!duplicated(rownames(count_data)), ]
head(count_data)
```
```{r}
library(readxl)
phenodata <- read_excel("C:/Users/rahma/Documents/Breast cancer RNA-Seq/Metadata.xlsx")
phenodata
```
Confirm the column names in counts data tallies with the rownames in the phenodata
```{r}
rownames(phenodata) = phenodata$`SRA Accession number`
phenodata
```
```{r}
rownames(phenodata)
```
```{r}
colnames(count_data)
```
```{r}
colnames(count_data) == rownames(phenodata)
```
```{r}
phenodata$Subtype <- factor(phenodata$Subtype)
class(phenodata$Subtype)
phenodata$Subtype
```
```{r}
phenodata$Subtype = relevel(phenodata$Subtype, ref = "Normal")
phenodata$Subtype
```

# Creating a DESeq dataset
```{r}
library(DESeq2)
library(genefilter)
```
```{r}
dds <- DESeqDataSetFromMatrix(countData = count_data, colData = phenodata, design = ~Subtype)
dds
```
```{r}
smallestGroupSize <- 3
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds <- dds[keep,]
dds
```

# Exploratory data analysis
## Estimate sequencing depth
```{r}
dds <- estimateSizeFactors(dds)
plot(sizeFactors(dds), colSums(counts(dds)))
abline(lm(colSums(counts(dds))~ sizeFactors(dds)))
```

## Data transformation and Variance stabilization
```{r}
rls <- rlog(dds)
plot(assay(rls)[,1], assay(rls)[,2], cex= .1)
```
```{r}
library(vsn)
meanSdPlot(assay(rls), main ="rlog transformation", ranks = FALSE)
```
```{r}
plotPCA(rls, intgroup = "Subtype")
```
```{r}
plotPCA(rls, intgroup= "Location")
```

## Hierarchical clustering
```{r}
plot(hclust(dist(t(assay(rls)))), labels= colData(rls)$Subtype)
```
# There might be some batch effects in the data. 
```{r}
library(sva)
idx= rowMeans(counts(dds, normalized= TRUE)) >1
dat <- counts(dds, normalized= TRUE)[idx,]
mod<- model.matrix(~ Subtype, colData(dds))
mod0 <- model.matrix(~1, colData(dds))

svBC <- svaseq(dat, mod, mod0, n.sv = 2)
dds$Location <- as.factor(dds$Location)
plot(svBC$sv[,1], svBC$sv[,2], col= dds$Location)

```
```{r}
dds.sva <- dds
dds.sva$SV1 <- svBC$sv[,1]
dds.sva$SV2 <- svBC$sv[,2]
design(dds.sva) <- ~ SV1 + SV2 + Subtype
```

# Differential gene expression analysis
```{r}
dds1 <- DESeq(dds.sva)
```
```{r}
resultsNames(dds1)
```

```{r}
TNBCvsNorm= results(dds1)
TNBCvsNorm
write.csv(TNBCvsNorm, file= "TNBC.csv")

```
```{r}
TNBCvsNorm= na.omit(TNBCvsNorm)
TNBC= TNBCvsNorm[abs(TNBCvsNorm$log2FoldChange) > 1 & TNBCvsNorm$padj < 0.01,]
TNBC
```

### Collecting the gene IDs from org.Hs.eg.db

```{r}
library(org.Hs.eg.db)
library(AnnotationDbi)
keys= rownames(TNBC)
symbolsTNBC= AnnotationDbi::select(org.Hs.eg.db, keys = keys, keytype = "ENTREZID", columns = c("SYMBOL", "GENENAME"))
head(symbolsTNBC)
```

```{r}
TNBC= cbind(TNBC, symbolsTNBC$SYMBOL, symbolsTNBC$GENENAME)
TNBC = data.frame(TNBC)
head(TNBC)
```

```{r}
upregTNBC= TNBC[TNBC$log2FoldChange >1,]
head(upregTNBC)
```
```{r}
downregTNBC <-TNBC[TNBC$log2FoldChange <1,]
head(downregTNBC)

```
```{r}
library(EnhancedVolcano)
EnhancedVolcano(TNBC, lab= TNBC$symbolsTNBC.SYMBOL, x="log2FoldChange", y= 'padj', title= "Normal vs TNBC")
```
```{r}
resSort <- TNBC[order(TNBC$padj),]
library(pheatmap)
topgenesTNBC= head(rownames(resSort), 20)
matTNBC <- assay(rls)[, 9:16][topgenesTNBC,]
matTNBC <- matTNBC - rowMeans(matTNBC)
rownames(matTNBC) <- symbolsTNBC$SYMBOL[match(rownames(matTNBC), symbolsTNBC$ENTREZID)]
pheatmap(mat = matTNBC)
```

```{r}
resNorm_ERPR <- results(dds1, name="Subtype_ERPRpos_vs_Normal")
resNorm_ERPR
write.csv(resNorm_ERPR, file = "ERPR.csv")
```
```{r}
resNorm_ERPR = na.omit(resNorm_ERPR)
ERPR = resNorm_ERPR[abs(resNorm_ERPR$log2FoldChange) >= 1 & resNorm_ERPR$padj < 0.01,]
ERPR
```
```{r}
keys= rownames(ERPR)
symbolERPR= AnnotationDbi::select(org.Hs.eg.db, keys = keys, keytype = "ENTREZID", columns = c("SYMBOL", "GENENAME"))
head(symbolERPR)
```

```{r}
ERPR = cbind(ERPR, symbolERPR$SYMBOL, symbolERPR$GENENAME)
head(ERPR)
```
```{r}
upregERPR = ERPR[ERPR$log2FoldChange > 1,]
upregERPR= upregERPR[order(upregERPR$log2FoldChange, decreasing = TRUE),]
head(upregERPR)
```

```{r}
downregERPR = ERPR[ERPR$log2FoldChange < 1,]
downregERPR= downregERPR[order(downregERPR$log2FoldChange, decreasing = FALSE),]
head(downregERPR)
```
```{r}
EnhancedVolcano(ERPR, lab = ERPR$`symbolERPR$SYMBOL`, x= "log2FoldChange", y= "padj")
```
```{r}
resSortERPR <- ERPR[order(ERPR$padj),]
library(pheatmap)
topgenesERPR= head(rownames(resSortERPR), 20)
matERPR <- assay(rls)[, c(5:8, 13:16)][topgenesERPR,]
matERPR <- matERPR - rowMeans(matERPR)
rownames(matERPR) <- symbolERPR$SYMBOL[match(rownames(matERPR), symbolERPR$ENTREZID)]
pheatmap(mat = matERPR)
```


```{r}
resNorm_ER <- results(dds1, name= "Subtype_ERpos_vs_Normal")
resNorm_ER
write.csv(resNorm_ER, file = "ER.csv")
```
```{r}
resNorm_ER= na.omit(resNorm_ER)
ER = resNorm_ER[abs(resNorm_ER$log2FoldChange) > 1 & resNorm_ER$padj < 0.01 ,]
ER
```
```{r}
keys = rownames(ER)
symbolER = AnnotationDbi::select(org.Hs.eg.db, keys = keys, keytype = "ENTREZID", columns = c("SYMBOL", "GENENAME"))
head(symbolER)
```

```{r}
ER = cbind(ER, symbolER$SYMBOL, symbolER$GENENAME)
ER
```
```{r}
upregER = ER[ER$log2FoldChange > 1,]
upregER = upregER[order(upregER$log2FoldChange, decreasing = TRUE),]
head(upregER)
```
```{r}
downregER = ER[ER$log2FoldChange < 1,]
downregER = downregER[order(downregER$log2FoldChange, decreasing = F),]
head(downregER)
```
```{r}
EnhancedVolcano(ER, lab = ER$`symbolER$SYMBOL`, x= "log2FoldChange", y= "padj")
```
```{r}
resSortER <- ER[order(ER$padj),]
topgenesER= head(rownames(resSortER), 20)
matER <- assay(rls)[, c(1:4, 13:16)][topgenesER,]
matER <- matER - rowMeans(matER)
rownames(matER) <- symbolER$SYMBOL[match(rownames(matER), symbolER$ENTREZID)]
pheatmap(mat = matER)
```

```{r}
resNorm_HER <- results(dds1, name="Subtype_HER2pos_vs_Normal")
resNorm_HER
write.csv(resNorm_HER, file = "HER2.csv")
```
```{r}
resNorm_HER= na.omit(resNorm_HER)
HER = resNorm_HER[abs(resNorm_HER$log2FoldChange) > 1 & resNorm_HER$padj < 0.01,]
HER
```
```{r}
keys= rownames(HER)
symbolHER = AnnotationDbi::select(org.Hs.eg.db, keys = keys, keytype = "ENTREZID", columns = c("SYMBOL", "GENENAME"))
head(symbolHER)
```
```{r}
HER = cbind(HER, symbolHER$SYMBOL, symbolHER$GENENAME)
HER
```
```{r}
upregHER= HER[HER$log2FoldChange > 1,]
upregHER = upregHER[order(upregHER$log2FoldChange, decreasing = TRUE),]
head(upregHER)
```
```{r}
downregHER= HER[HER$log2FoldChange < 1,]
downregHER = downregHER[order(downregHER$log2FoldChange, decreasing = F),]
head(downregHER)
```
```{r}
EnhancedVolcano(HER, lab= HER$`symbolHER$SYMBOL`, x= "log2FoldChange", y= "padj")
```
```{r}
resSortHER <- HER[order(HER$padj),]
topgenesHER= head(rownames(resSortHER), 20)
matHER <- assay(rls)[, 13:20][topgenesHER,]
matHER <- matHER - rowMeans(matHER)
rownames(matHER) <- symbolHER$SYMBOL[match(rownames(matHER), symbolHER$ENTREZID)]
pheatmap(mat = matHER)
```

```{r}
sessionInfo()
```





