---
title: "Gene and Pathway Enrichment Analysis"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

## IMPORT NEEDED PACKAGES

```{r}
library(clusterProfiler)
library(enrichplot)
# we use ggplot2 to add x axis labels (ex: ridgeplot)
library(ggplot2)
library(org.Hs.eg.db)
```
# Triple Negative Breast Cancer
## Import TNBC

```{r}
TNBC_df= read.csv("TNBC.csv", header = TRUE)
head(TNBC_df)
```
```{r}
TNBCgene_list = TNBC_df$log2FoldChange

#name the vector
names(TNBCgene_list) <- TNBC_df$X

head(TNBCgene_list)
```
```{r}
TNBCgene_list <- na.omit(TNBCgene_list)
TNBCgene_list <- sort(TNBCgene_list, decreasing = TRUE)
head(TNBCgene_list)
```
## Overrepresentaion analysis
```{r}
enrichedTNBC <- enrichGO(gene = names(TNBCgene_list),
                         OrgDb = org.Hs.eg.db,
                         keyType = "ENTREZID",
                         ont = "ALL",
                         pAdjustMethod = "BH")
head(enrichedTNBC)
write.csv(enrichedTNBC, "ORAGOTNBC.csv")
```
Visualizing enrichment results
```{r}
# First construct similarity matrix
egoTNBC <- pairwise_termsim(enrichedTNBC)
emapplot(egoTNBC)
```
## Gene Set Enrichment analysis
```{r}
gseTNBC <- gseGO(geneList=TNBCgene_list, 
             ont ="ALL", 
             keyType = "ENTREZID", 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = org.Hs.eg.db, 
             pAdjustMethod = "BH")
write.csv(gseTNBC, "GeneEnrichmentTNBC.csv")
```
```{r}
head(gseTNBC)
```
```{r}
dotplot(gseTNBC, showCategory=10, split=".sign") + facet_grid(.~.sign)
```
Enrichment map for gene set enrichment analysis
```{r}
# Compute term similarity matrix
egseaTNBC <- pairwise_termsim(gseTNBC)

# Visualize results
emapplot(egseaTNBC, showCategory = 30)
```

```{r}
gseaplot(gseTNBC, by = "all", title = gseTNBC$Description[1], geneSetID = 1)
```

```{r}
gseaplot(gseTNBC, by = "all", title = gseTNBC$Description[2], geneSetID = 2)
```

# KEGG Analysis
## KEGG ORA

```{r}
keggORATNBC <- enrichKEGG(names(TNBCgene_list),
                          organism = 'hsa',
                          keyType = "ncbi-geneid")
head(keggORATNBC)
write.csv(keggORATNBC, "ORAKEGGTNBC.csv")
```

```{r}
eKEGG <- pairwise_termsim(keggORATNBC)
emapplot(eKEGG)
```

```{r}
kk2 <- gseKEGG(TNBCgene_list, organism = "hsa", keyType = "ncbi-geneid", pAdjustMethod = "BH")
head(kk2)
write.csv(kk2, "PathwayEnrichmentTNBC.csv")
```
```{r}
dotplot(kk2, showCategory = 15, title= "Enriched pathways", split=".sign") + facet_grid(.~.sign)
write.csv(kk2, "TNBC KEGG.csv")
```


```{r}
kk2_ema <- pairwise_termsim(kk2)
emapplot(kk2_ema)
```



```{r}
ridgeplot(kk2, showCategory = 15)+ labs( x= "Enrichment distribution")
```


# Estrogen receptor positive BC
```{r}
ER_df <- read.csv("ER.csv", header = TRUE)
head(ER_df)
```
```{r}
ERgene_list <- ER_df$log2FoldChange
names(ERgene_list) <- ER_df$X
ERgene_list <- na.omit(ERgene_list)
ERgene_list <- sort(ERgene_list, decreasing= TRUE)
```

## Gene Overrepresentation analysis
```{r}
ERORA <- enrichGO(gene = names(ERgene_list),
                  OrgDb = org.Hs.eg.db,
                  keyType = "ENTREZID",
                  ont = "ALL")
head(ERORA)
write.csv(ERORA, "ORAGOEP.csv")
```
```{r}
egoER <- pairwise_termsim(ERORA)
emapplot(egoER)
```

## Gene set enrichment analysis
```{r}
gseaER <- gseGO(geneList <- ERgene_list, 
                ont = "ALL",
                OrgDb = org.Hs.eg.db)
write.csv(gseaER, "ER GO.csv")
head(gseaER)
write.csv(gseaER, "GeneEnrichmentER.csv")
```
```{r}
gseGOER <- pairwise_termsim(gseaER)
emapplot(gseGOER)
```
```{r}  
dotplot(gseGOER, showCategory = 10, split= ".sign") + facet_grid(.~.sign)
```

## KEGG pathway enrichment analysis
## ORA
```{r}
ORAERKegg <- enrichKEGG(gene = names(ERgene_list),
                        organism = "hsa",
                        keyType = "ncbi-geneid")
head(ORAERKegg)
write.csv(ORAERKegg, "ORAKEGGER.csv")
```
```{r}
eKEGGER <- pairwise_termsim(ORAERKegg)
emapplot(eKEGGER)
```

KEGG Enrichment Analysis
```{r}
enrKEGGER <- gseKEGG(ERgene_list, keyType = "ncbi-geneid")
write.csv(enrKEGGER, "ER KEGG.csv")
head(enrKEGGER) 
write.csv(enrKEGGER, "PathwayEnrichmentER.csv")
```
```{r}
dotplot(enrKEGGER, showCategory = 10, split= ".sign") + facet_grid(.~.sign)
```
```{r}
ridgeplot(enrKEGGER, showCategory = 15)
```


# HER2+ breast cancer
## ORA
```{r}
HER2_df <- read.csv("HER2.csv", header = TRUE)
head(HER2_df)
```
```{r}
HERgene_list <- HER2_df$log2FoldChange
names(HERgene_list) <- HER2_df$X
HERgene_list <- na.omit(HERgene_list)
HERgene_list <- sort(HERgene_list, decreasing = TRUE)
```

## Overrepresentation Analysis
```{r}
library(org.Hs.eg.db)
library(clusterProfiler)
egoHER <- enrichGO(names(HERgene_list),
                   OrgDb = org.Hs.eg.db,
                   ont = "ALL",
                   keyType = "ENTREZID")
head(egoHER)
write.csv(egoHER, "ORAGOHER.csv")
```
```{r}
eoRAHER <- pairwise_termsim(egoHER)
emapplot(eoRAHER)
```
```{r}
HERgSE <- gseGO(geneList = HERgene_list,
                ont = "ALL",
                OrgDb = org.Hs.eg.db)
write.csv(HERgSE, "GeneEnrichmentHER.csv")
```

```{r}
head(HERgSE)
```
```{r}
dotplot(HERgSE, showCategory= 10, split= ".sign") + facet_grid(.~.sign)
```

## KEGG Enrichment
### ORA
```{r}
HERKEGGORA <- enrichKEGG(names(HERgene_list), organism = "hsa", keyType = "ncbi-geneid")
head(HERKEGGORA)
write.csv(HERKEGGORA, "ORAKEGGHER.csv")
```

```{r}
eKeggHER <- pairwise_termsim(HERKEGGORA)
emapplot(eKeggHER)
```

KEGG Gene enrichment
```{r}
gseHERKEGG <- gseKEGG(HERgene_list, keyType = "ncbi-geneid")
write.csv(gseHERKEGG, "PathwayEnrichmentHER.csv")
head(gseHERKEGG)
```
```{r}
dotplot(gseHERKEGG, showCategory  = 10, split = ".sign") + facet_grid(.~.sign)
```

```{r}
ridgeplot(gseHERKEGG, showCategory = 15)
```



```{r}
sessionInfo()
```

